The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-11 rational.tex
sec fractional types

%%%%%

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{1/b}}. These two values are free to move independently
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

\paragraph*{Entanglement and Anti-Particles.}
A dual view of the operation of {{eta*}} and {{eps*}} is inspired by
quantum mechanics. The {{eta*}} is a site of fission that creates an
information particle and its anti-particle. Both flow in the same
direction in time. The produce particles are entangles in that actions
on one (such as transformation by application of an isomorphism)
affects the other. Further the fission site does not create a particle
in any specific state, but in a superposition of all possible states
as dictated by its type. In each possible world that the particle
exists, it takes on a specific value inhibiting its type and
correspondingly its entangled anti-particle takes the dual of the
specific value.

This also leads to the exciting point of view that entangled
particle/anti-particle pairs act as functions in a physical
first-class sense. We can ask the function to be applied to a value by
unifying the anti-particle with a particle in a well-known state and
examining is pair. We shall use this point of view in the development
our SAT solver (see Sec. \ref{sec:sat-solver}).

\paragraph*{Coherence for Compact Closure with Multiplicative Inverses} 
As before, the requisite coherence condition on Compact Closed
Categories is applicable and is expressed below in terms of
{{eta*}}/{{eps*}} and the multiplicative monoid.

\paragraph*{Annihilation.}
As with {{(+, 0)}} we can construct a {{trace*}} operation. Unlike the
additive {{trace+}}, {{trace*}} gives us the ability to find the
fixpoint of the constraint (expressed with combinators) that is
traced. If the constraint has not satisfying values, there is not
fixpoint possible -- we use the term `annihilation' to describe the
corresponding undefined state of the program. Annihilation and a
non-termination both describe undefined program execution -- while
non-termination characterizes an iteration that fails to terminate,
annihilation characterizes a constraint that has no fixpoint.

>>>>

This can be understood in terms of the \emph{logic programming} as
follows. Operator {{eta* : 1 <-> (1/b)*b}} acts as the site of
creation of a new typed logic variable and constructs from it a value
of type {{b}} and its dual of type {{1/b}}. These two values are free
to move independently through the circuit. Operations on one value
affect what operations are possible on the other since they are both
related by an underlying logic variable. Dually, {{eps*}} acts as a
unification site for values of type {{b}} and {{1/b}} flowing along
both branches. If the values are not unifiable duals, then the
specific program sequence fails. Like with logic programming, one can
has many worlds where some might fail, while others make
progress. Sec. \ref{sec:rat} has more details.  If the values are
exactly duals, then unification succeeds and {{eta*}} returns {{1}}.

% \paragraph*{Entanglement and Anti-Particles.}

An analogy inspired by quantum mechanics is also applicable here.  The
operator {{eta*}} is a site of fission that creates \emph{an entangled
  particle and anti-particle}. These particles can be thought to be in
a \emph{superposition} of states determined by their type. Both flow
in the same direction in time.  In each possible world that the
particle exists, it takes on a specific value inhibiting its type and
correspondingly its entangled anti-particle takes the dual of the
specific value. Since the particles are entangled, actions on one
(such as transformation by application of an isomorphism) affects the
other in much the same sense as \emph{action at a distance}. The
`{{a-o*b}}' functions mentioned in the introduction, are first-class
values and they correspond to these entagled pairs. Finally {{eps*}}
operations are sites where corresponding particle and anti-particles
annihilate each other. While appealing as an analogy, this description
does not imply (or preclude) any formal connection with Physics.

%% We can ask the function to be applied to a value by
%% unifying the anti-particle with a particle in a well-known state and
%% examining is pair. We shall use this point of view in the development
%% our SAT solver (see Sec. \ref{sec:sat-solver}).

%%  This also leads to the exciting point of view that
%% entangled such particle/anti-particle pairs, which are first class
%% values, act as functions in a physical first-class sense.

%% Further the fission site does not create a particle in any specific
%% state, but in a superposition of all possible states as dictated by
%% its type.

% Since such pairs are first-class entities, they correspond to

\emph{Coherence for Compact Closure with Multiplicative Duality.}  As
before, the triangle axiom which is the coherence condition on compact
closed categories is applicable and is expressed below in terms of
{{eta*}}/{{eps*}} and the multiplicative monoid.
\begin{center}
  \includegraphics{diagrams/coherence.pdf}
\end{center}

\emph{Annihilation.}  
With {{eta*}} and {{eps*}}, we can construct a {{trace*}} operation
which gives us the ability to find the fixpoint of a constraint.  If
the constraint has no satisfying values, there is no fixpoint possible
-- we use the term \emph{annihilation} to describe the corresponding
undefined state of the program.

Annihilation and non-termination describe undefined program
execution. While non-termination characterizes an iteration that fails
to terminate, annihilation characterizes a constraint that has no
fixpoint. The SAT-solver of Sec. \ref{sec:sat-solver} works by
annihilating only the program states that do not match the boolean
satisfiability constraint.
